---
title: "2 Workflow: basics"
format: pdf
editor: visual
---

### **2.1 Coding basics**

### Basic math calculations

```{r}

1 / 200 * 30

(59 + 73 + 2) / 3

sin(pi / 2)
```

### Create new objects

```{r}

x <- 3 * 4

x
```

### Combine multiple elements into a vector

```{r}

primes <- 
  c(2, 3, 5, 7, 11, 13)
```

### Basic arithmetic on vectors is applied to every element of the vector

```{r}

primes * 2

primes - 1
```

### Create object

```{r}

# object_name <- value
x <- 5
```

### **2.2 Comments**

```{r}

# create vector of primes
primes <- 
  c(2, 3, 5, 7, 11, 13)

# multiply primes by 2
primes * 2
```

### **2.4 Calling functions**

```{r}

seq(from = 1, to = 10)

seq(1, 10)
```

### **2.5 Exercises**

1.  Why does this code not work? - "i"

```{r}
#| eval: false

my_variable <- 10

my_varıable
```

2.  Tweak each of the following R commands so that they run correctly:

```{r}
#| eval: false

libary(todyverse)

ggplot(dTA = mpg) + 
  geom_point(maping = aes(x = displ y = hwy)) +
  geom_smooth(method = "lm)
```

```{r}
#| message: false
#| warning: false

library(tidyverse)

ggplot(data = mpg) + 
  geom_point(
    mapping = aes(
      x = displ, 
      y = hwy
    )
  ) +
  geom_smooth(
    mapping = aes(
      x = displ, 
      y = hwy
    ),
    method = "lm"
  )
```

3.  Press Option + Shift + K / Alt + Shift + K. What happens? How can you get to the same place using the menus?

    Help \> Keyboard Shortcuts Help

4.  Let’s revisit an exercise from the [Section 1.6](https://r4ds.hadley.nz/data-visualize.html#sec-ggsave). Run the following lines of code. Which of the two plots is saved as `mpg-plot.png`? Why?

    First plot is saved, because it is specified in the arguments of ggsave

```{r}
#| message: false
#| warning: false

my_bar_plot <- 
  ggplot(
    mpg, 
    aes(x = class)
  ) +
  geom_bar()

my_scatter_plot <- 
  ggplot(
    mpg, 
    aes(x = cty, y = hwy)
  ) +
  geom_point()

ggsave(
  filename = "mpg-plot.png", 
  plot = my_bar_plot
)
```
