---
title: "3 Data transformation"
format: html
editor: visual
---

### **3.1 Introduction**

Prerequisites

```{r}
#| message: false
#| warning: false

library(nycflights13)
library(tidyverse)

```

#### nycflights13

This dataset contains all 336,776 flights that departed from New York City in 2013. The data comes from the US Bureau of Transportation Statistics and is documented in ?flights.

```{r}

flights

```

#### tibbles

flights is a tibble. Tibbles cannot access a column when you provide a partial name of the column, but data frames can.

With tibble, recycling only happens for a length 1 input.

- https://tibble.tidyverse.org/reference/tbl_df-class.html

- vignette("tibble")

- https://r4ds.had.co.nz/tibbles.html

```{r}
#| results: 'hide'

flights$yea

df_flights <- as.data.frame(flights)

df_flights$yea 

```

When you access only one column of a tibble, it will keep the tibble structure. But when you access one column of a data frame, it will become a vector.

| Expression            | Tibble returns  | Data frame returns  |
|-----------------------|-----------------|---------------------|
| flights\["year"\]     | 1-column tibble | 1-column data frame |
| flights\[, "year"\]   | 1-column tibble | vector              |
| flights\[\["year"\]\] | vector          | vector              |
| flights\$year         | vector          | vector              |

```{r}

flights[, "year"]
is_tibble(flights[, "year"])

```

```{r}
#| results: 'hide'

df_flights[, "year"] # will print out all rows

```

```{r}

is.data.frame(df_flights[, "year"])
is.vector(df_flights[, "year"])

```

However, other forms of subsetting, including \[\[ \]\] and \$, work the same for tibbles and data frames.

```{r}
#| results: 'hide'

flights[["year"]] # will print out all rows
df_flights[["year"]] # will print out all rows

```

```{r}

is_tibble(flights[["year"]])
is.data.frame(df_flights[["year"]])

is.vector(flights[["year"]])
is.vector(df_flights[["year"]])

```

#### Other ways to look at the data.

#### glimpse():

- <int> is short for integer
- <dbl> is short for double (aka real numbers)
- <chr> for character (aka strings)
- <dttm> for date-time.

```{r}

View(flights)
print(flights, width = Inf)
glimpse(flights)
summary(flights)

```

#### dplyr basics

- The first argument is always a data frame

- The subsequent arguments typically describe which columns to operate on using the variable names (without quotes)

- The output is always a new data frame

dplyr’s verbs are organized into four groups based on what they operate on: rows, columns, groups, or tables.

```{r}

flights |>
  filter(dest == "IAH") |> 
  group_by(year, month, day) |> 
  summarize(
    arr_delay = mean(arr_delay, na.rm = TRUE)
  )

```

### **3.2 Rows**

#### filter()

- \> (greater than)

- \>= (greater than or equal to)

- \< (less than)

- \<= (less than or equal to)

- == (equal to)

- != (not equal to)

- combine conditions with & or , to indicate “and” (check for both conditions)

- combine conditions with \| to indicate “or” (check for either condition)

Find all flights that departed more than 120 minutes (two hours) late:

```{r}

flights |> 
  filter(dep_delay > 120)

```

Flights that departed on January 1:

```{r}

flights |> 
  filter(month == 1 & day == 1)

```

Flights that departed in January or February:

```{r}

flights |> 
  filter(month == 1 | month == 2)

```

combine \| and == with %in%. It keeps rows where the variable equals one of the values on the right:

```{r}

flights |> 
  filter(month %in% c(1, 2))

```

#### arrange()

If you provide more than one column name, each additional column will be used to break ties in the values of the preceding columns. We get the earliest years first, then within a year, the earliest months, etc.

```{r}

flights |> 
  arrange(year, month, day, dep_time)

```

desc() Use inside of arrange() to re-order the data frame based on that column in descending (big-to-small) order

```{r}

flights |> 
  arrange(desc(dep_delay))

```

#### distinct()

Finds all the unique rows in a dataset, so technically, it primarily operates on the rows. Most of the time, however, you’ll want the distinct combination of some variables, so you can also optionally supply column names.

Remove duplicate rows, if any:

```{r}

flights |> 
  distinct()

```

Find all unique origin and destination pairs:

```{r}

flights |> 
  distinct(origin, dest)

```

Alternatively, if you want to keep the other columns when filtering for unique rows, you can use the .keep_all = TRUE option.

It’s not a coincidence that all of these distinct flights are on January 1: distinct() will find the first occurrence of a unique row in the dataset and discard the rest.

```{r}

flights |> 
  distinct(origin, dest, .keep_all = TRUE)

```

#### count()

Find the number of occurrences. With the sort = TRUE argument, you can arrange them in descending order of the number of occurrences.

```{r}

flights |>
  count(origin, dest, sort = TRUE)

```

### **Exercises**

1.  In a single pipeline for each condition, find all flights that meet the condition:

- Had an arrival delay of two or more hours

```{r}

flights |>
  filter(arr_delay >= 120)
  
```

- Flew to Houston (IAH or HOU)

```{r}

flights |>
  filter(dest %in% c("IAH", "HOU"))

```

- Were operated by United, American, or Delta

```{r}

flights |>
  filter(carrier %in% c("UA", "AA", "DL"))

```

- Departed in summer (July, August, and September)

```{r}

flights |>
  filter(month %in% c(7, 8, 9))

```

- Arrived more than two hours late but didn’t leave late

```{r}

flights |>
  filter(arr_delay > 120 & dep_delay <= 0)

```

- Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}

flights |>
  filter(dep_delay >= 60, dep_delay - arr_delay > 30)

```

2.  Sort flights to find the flights with the longest departure delays. Find the flights that left earliest in the morning.

```{r}

# longest departure delays
flights |>
  arrange(desc(dep_delay))

```

```{r}

# left earliest in the morning
flights |>
  arrange(sched_dep_time) |>
  relocate(sched_dep_time)

```

3.  Sort flights to find the fastest flights. (Hint: Try including a math calculation inside of your function.)

```{r}

flights |>
  arrange(desc(distance / (air_time * 60)))
  
```

4.  Was there a flight on every day of 2013?

Yes

```{r}

flights |> 
  distinct(year, month, day) |>
  nrow()

flights |> 
  distinct(year, month, day) |>
  count()
```

5.  Which flights traveled the farthest distance? Which traveled the least distance?

```{r}

# farthest distance
flights |>
  arrange(desc(distance)) |>
  head(10) |>
  relocate(distance)

```

```{r}

# least distance
flights |>
  arrange(distance) |>
  head(10) |>
  relocate(distance)

```

6.  Does it matter what order you used filter() and arrange() if you’re using both? Why/why not? Think about the results and how much work the functions would have to do.

Regarding results - it doesn't matter. Regarding performance - it does.

filter() reduces the number of rows. arrange() must sort all rows it receives.

- filter() then arrange():

filter() removes many rows first \> arrange() sorts only the remaining (smaller) set Less work -\> faster, especially on large datasets.

- arrange() then filter():

arrange() sorts the full dataset \> filter() then removes rows. More work -\> slower, because sorting is done on all rows.

So for efficiency, generally filter first, then arrange.

### **3.3 Columns**

There are four important verbs that affect the columns without changing the rows: - mutate() creates new columns that are derived from the existing columns - select() changes which columns are present - rename() changes the names of the columns - relocate() changes the positions of the columns

### mutate()

Compute the gain, how much time a delayed flight made up in the air, and the speed in miles per hour

```{r}

flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  ) |>
  relocate(gain, speed)

```

### .before

By default, mutate() adds new columns on the right-hand side of your dataset. Use the .before argument adds the variables to the left-hand side

```{r}

flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .before = 1
  )

```

### .after

Use .after to add after a variable, and in both .before and .after you can use the variable name instead of a position. For example, we could add the new variables after day:

```{r}

flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .after = day
  )

```

### .keep

Control which variables are kept with the .keep argument. A particularly useful argument is "used" - only keeps the columns that were involved or created in the mutate() step.

For example, the following output will contain only the variables dep_delay, arr_delay, air_time, gain, hours, and gain_per_hour.

```{r}

flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )

```

### select()

Select columns by name:

```{r}

flights |> 
  select(year, month, day)

```

Select all columns between year and day (inclusive):

```{r}

flights |> 
  select(year:day)

```

Select all columns except those from year to day (inclusive):

```{r}

flights |> 
  select(!year:day)

```

Select all columns that are characters:

```{r}

flights |> 
  select(where(is.character))

```

There are a number of helper functions you can use within select().

https://tidyselect.r-lib.org/reference/language.html

- starts_with("abc"): matches names that begin with “abc”:

```{r}

flights |> 
  select(starts_with("dep"))

```

- ends_with("xyz"): matches names that end with “xyz”:

```{r}

flights |> 
  select(ends_with("time"))

```

- contains("ijk"): matches names that contain “ijk”:

```{r}

flights |> 
  select(contains("dep"))

```

- num_range("x", 1:3): matches x1, x2 and x3 (pick columns with a common prefix plus numbers):

```{r}

flights |> 
  select(num_range("dep", 1:2))

```

or matches(): matches a regular expression.

Rename variables as you select() them by using =

```{r}

flights |> 
  select(tail_num = tailnum)

```

### rename()

If you want to keep all the existing variables and just want to rename a few:

```{r}

flights |> 
  rename(tail_num = tailnum)

```

### janitor::clean_names()

If you have a bunch of inconsistently named columns and it would be painful to fix them all by hand. Resulting names are unique and consist only of the \_ character, numbers, and letters. Capitalization preferences can be specified using the case parameter.

### relocate()

By default the variables are moved to the front:

```{r}

flights |> 
  relocate(time_hour, air_time)

```

You can also specify where to put them using the .before and .after arguments:

```{r}

# .after
flights |> 
  relocate(year:dep_time, .after = time_hour)

```

```{r}

# .before
flights |> 
  relocate(starts_with("arr"), .before = dep_time)

```

### Exercises

1.  Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?

One would expect dep_delay to equal the difference between dep_time and sched_dep_time:

dep_delay = dep_time − sched_dep_time

However, the times are in HHMM format (e.g., 1355 = 13:55), not continuous minutes, so they can't be directly subtracted:

```{r}

flights |>
  select(dep_time, sched_dep_time, dep_delay) |>
  mutate(
    diff = dep_time - sched_dep_time, 
    match = dep_delay == diff
  ) |>
  count(match)
```

554 - 600 = -46 doesn't match dep_delay (e.g., -6) because the HHMM format doesn't represent continuous time

```{r}
flights |>
  select(dep_time, sched_dep_time, dep_delay) |>
  mutate(
    diff = dep_time - sched_dep_time, 
    match = dep_delay == diff
  ) |>
  filter(match == FALSE)
```

Solution

```{r}

flights |> 
  mutate(
    # time_hour is scheduled departure as POSIXct (includes year, month, day, hour)
    # dep_delay is in minutes, so multiply by 60 to get seconds
    actual_dep_ts = time_hour + as.duration(dep_delay * 60),
    # Create scheduled dep_ts from sched_dep_time + time_hour's date
    sched_hours = sched_dep_time %/% 100,
    sched_mins = sched_dep_time %% 100,
    sched_dep_ts = time_hour - as.duration((sched_hours * 60 + sched_mins) * 60) + as.duration(sched_hours * 3600 + sched_mins * 60)
  ) |> 
  mutate(
    calculated_delay = as.numeric(
      actual_dep_ts - sched_dep_ts, units = "mins"
    ), 
    match = calculated_delay == dep_delay
  ) |> 
  select(dep_time, sched_dep_time, dep_delay, calculated_delay, match) |> 
  filter(!is.na(dep_delay) & !is.na(calculated_delay)) #|> 
  #count(match)

```

2.  Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

```{r}

flights |> 
  select(dep_time, dep_delay, arr_time, arr_delay)

```

```{r}

flights |> 
  select(starts_with("dep") | starts_with("arr"))

```

```{r}

flights |> 
  select(
    (starts_with("dep") | starts_with("arr"))
    & (ends_with( "time") | ends_with("delay"))
  )
```

```{r}

flights |>
  select(
    (ends_with("delay") | ends_with("time")) 
    & !starts_with("sched")  
    & !starts_with("air")
  )

```

```{r}

flights |>
  select(
    (contains("dep") | contains("arr")) 
    & -contains("sched")
  )

```

3.  What happens if you specify the name of the same variable multiple times in a select() call?

Variable is selected only once.

```{r}

flights |>
  select(arr_time, arr_time)
```

4.  What does the any_of() function do? Why might it be helpful in conjunction with this vector?

Selects only the variables that exist in a data frame from a given character vector, without throwing an error if some variables are missing https://tidyselect.r-lib.org/reference/all_of.html

```{r}

variables <- 
  c("year", "month", "day", "dep_delay", "arr_delay")

flights |>
  select(any_of(variables))

```

5.  Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?

All tidyselect select helpers have ignore.case = TRUE by default, meaning they ignore uppercase/lowercase differences. The case-insensitive default can lead to unexpected matches (e.g., contains("ID") matches "user_id" and "ID_2").

https://tidyselect.r-lib.org/reference/starts_with.html https://r-statistics.co/dplyr-contains-in-R.html

```{r}

flights |> 
  select(contains("TIME"))

```

```{r}

# case sencitive
flights |> 
  select(contains("TIME", ignore.case = FALSE))

```

6.  Rename air_time to air_time_min to indicate units of measurement and move it to the beginning of the data frame.

```{r}

flights |>
  rename(air_time_min = air_time) |>
  relocate(air_time_min)

```

7.  Why doesn’t the following work, and what does the error mean?

Because only tailnum is selected. arr_delay doesn't exist anymore by the time arrange() is executed.

```{r}

flights |> 
  select(tailnum) |> 
  arrange(arr_delay)

```

### **3.4 The pipe**

```{r}

flights |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))

```

### **3.5 Groups**

### group_by()

Use group_by() to divide your dataset into groups meaningful for your analysis:

```{r}

flights |> 
  group_by(month)

```

group_by() doesn’t change the data but, if you look closely at the output, you’ll notice that the output indicates that it is “grouped by” month (Groups: month \[12\]). This means subsequent operations will now work “by month”. group_by() adds this grouped feature (referred to as class) to the data frame, which changes the behavior of the subsequent verbs applied to the data.

### summarize()

```{r}

# some of the observed flights have missing data in the delay column, and so when calculating the mean including those values, we got an NA result
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay)
  )

```

### na.rm

```{r}

# average departure delay by month
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE)
  )

```

### n()

Returns the number of rows in each group:

```{r}

flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE), 
    n = n()
  )

```

### The slice\_ functions

There are five handy functions that allow you to extract specific rows within each group:

- df \|\> slice_head(n = 1) takes the first row from each group
- df \|\> slice_tail(n = 1) takes the last row in each group
- df \|\> slice_min(x, n = 1) takes the row with the smallest value of column x
- df \|\> slice_max(x, n = 1) takes the row with the largest value of column x
- df \|\> slice_sample(n = 1) takes one random row

You can vary n to select more than one row, or instead of n =, you can use prop = 0.1 to select (e.g.) 10% of the rows in each group.

For example, the following code finds the flights that are most delayed upon arrival at each destination:

```{r}

flights |> 
  group_by(dest) |> 
  slice_max(arr_delay, n = 1) |>
  relocate(dest)

```

Note that there are 105 destinations but we get 108 rows here. What’s up? slice_min() and slice_max() keep tied values so n = 1 means give us all rows with the highest value. If you want exactly one row per group you can set with_ties = FALSE.

This is similar to computing the max delay with summarize(), but you get the whole corresponding row (or rows if there’s a tie) instead of the single summary statistic.

### Grouping by multiple variables

```{r}

daily <- 
  flights |>  
  group_by(year, month, day)

daily

```

When you summarize a tibble grouped by more than one variable, each summary peels off the last group.

```{r}

daily_flights <- 
  daily |> 
  summarize(n = n())

# summarise()` has grouped output by 'year', 'month'. You can override using the `.groups` argument.

daily_flights

```

```{r}

daily_flights <- daily |> 
  summarize(
    n = n(), 
    .groups = "drop_last"
  )

```

### group_by test

```{r}

tb <-
  tibble(
    var_1 = c("a", "a", "a", "a", "b", "b", "b", "b"), 
    var_2 = c("aa", "aa", "ab", "ab", "ba", "ba", "bb", "bb"), 
    var_3 = c(1, 2, 3, 1, 2, 3, 1, 2)
  )

tb |>
  group_by(var_1, var_2) |>
  summarize(mean = mean(var_3))

```

### ungroup()

You might also want to remove grouping from a data frame without using summarize(). You can do this with ungroup()

```{r}

daily |> 
  ungroup()

```

What happens when you summarize an ungrouped data frame?

You get a single row back because dplyr treats all the rows in an ungrouped data frame as belonging to one group.

```{r}

daily |> 
  ungroup() |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE), 
    flights = n()
  )

```

### .by

You can now also use the .by argument to group within a single operation:

```{r}

flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n(),
    .by = month
  )

```

Or if you want to group by multiple variables:

https://tidyverse.org/blog/2023/02/dplyr-1-1-0-per-operation-grouping/

```{r}

flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n(),
    .by = c(origin, dest)
  )

```

### Exercises

1.  Which carrier has the worst average delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights \|\> group_by(carrier, dest) \|\> summarize(n()))

EV has the worst average departure delay at 20 minutes among major carriers with substantial flight volume - 54173. F9 has 20.2 minutes but only 681 flights, so it's less reliable as a comparison.

```{r}

flights |>
  group_by(carrier) |>
  summarize(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    n_flights = n()
  ) |>
  arrange(desc(avg_arr_delay))

```

Bad airports vs. bad carriers.

If a carrier has high delays across multiple destinations, it's likely the carrier. If a carrier has high delays only at specific destinations, it's likely the airport.

```{r}

flights |>
  group_by(carrier, dest) |>
  summarize(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n_flights = n()
  ) |>
  arrange(carrier, desc(avg_delay)) |>
  print(n = Inf)

```

```{r}

flights |>
  group_by(dest, carrier) |>
  summarize(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n_flights = n()
  ) |>
  arrange(dest, desc(avg_delay)) |>
  print(n = Inf)

```

2.  Find the flights that are most delayed upon departure to each destination.

```{r}

flights |>
  group_by(dest) |>
  filter(dep_delay == max(dep_delay, na.rm = TRUE)) |>
  select(dest, carrier, flight, dep_delay, arr_delay, year, month, day) |>
  arrange(desc(dep_delay))

```

3.  How do delays vary over the course of the day? Illustrate your answer with a plot.

```{r}

delays <-
  flights |>
  group_by(hour) |>
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE))

  ggplot(
    delays, 
    aes(x = hour, y = avg_dep_delay)
  ) + 
  geom_col()

```

4.  What happens if you supply a negative n to slice_min() and friends?

A negative n in slice_min() means “keep all but n rows,” rather than selecting the smallest n rows.

For example, slice_min(x, n = -2) returns the group size minus 2 rows with the smallest values, so it effectively drops 2 rows from the end of the ranked result

```{r}

flights |> 
  group_by(dest) |> 
  slice_min(arr_delay, n = 1) |>
  relocate(dest, arr_delay)

```

```{r}

flights |> 
  group_by(dest) |> 
  slice_min(arr_delay, n = -1) |>
  relocate(dest, arr_delay)

```

5.  Explain what count() does in terms of the dplyr verbs you just learned. What does the sort argument to count() do?

Groups the data and counts the number of rows for each group.

- sort = FALSE (default): Output is in the order groups appear in the data
- sort = TRUE: Output is sorted by the count in descending order (largest groups at the top)

```{r}

flights |> 
  count(carrier) 

flights |> 
  count(carrier, sort = TRUE)

```

6.  Suppose we have the following tiny data frame:

```{r}

df <- 
  tibble(
    x = 1:5,
    y = c("a", "b", "a", "a", "b"),
    z = c("K", "K", "L", "L", "K")
  )

df

```

a.  Write down what you think the output will look like, then check if you were correct, and describe what group_by() does.

```{r}

df |>
  group_by(y)

```

b.  Write down what you think the output will look like, then check if you were correct, and describe what arrange() does. Also, comment on how it’s different from the group_by() in part (a).

```{r}

df |>
  arrange(y)

```

c.  Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.

```{r}

df |>
  group_by(y) |>
  summarize(mean_x = mean(x))

```

d.  Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. Then, comment on what the message says.

```{r}

df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))

```

e.  Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. How is the output different from the one in part (d)?

```{r}

df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x), .groups = "drop")

```

f.  Write down what you think the outputs will look like, then check if you were correct, and describe what each pipeline does. How are the outputs of the two pipelines different?

```{r}

df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))

df |>
  group_by(y, z) |>
  mutate(mean_x = mean(x))

```

### **3.6 Case study: aggregates and sample size**

Whenever you do any aggregation, it’s always a good idea to include a count (n()). That way, you can ensure that you’re not drawing conclusions based on very small amounts of data. We’ll demonstrate this with some baseball data from the Lahman package. Specifically, we will compare what proportion of times a player gets a hit (H) vs. the number of times they try to put the ball in play (AB):

```{r}

batters <- 
  Lahman::Batting |> 
  group_by(playerID) |> 
  summarize(
    performance = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    n = sum(AB, na.rm = TRUE)
  )

batters

```

When we plot the skill of the batter (measured by the batting average, performance) against the number of opportunities to hit the ball (measured by times at bat, n), you see two patterns:

1.  The variation in performance is larger among players with fewer at-bats. The shape of this plot is very characteristic: whenever you plot a mean (or other summary statistics) vs. group size, you’ll see that the variation decreases as the sample size increases4.

2.  There’s a positive correlation between skill (performance) and opportunities to hit the ball (n) because teams want to give their best batters the most opportunities to hit the ball.

```{r}

batters |> 
  filter(n > 100) |> 
  ggplot(aes(x = n, y = performance)) +
  geom_point(alpha = 1 / 10) + 
  geom_smooth(se = FALSE)

```

This also has important implications for ranking. If you naively sort on desc(performance), the people with the best batting averages are clearly the ones who tried to put the ball in play very few times and happened to get a hit, they’re not necessarily the most skilled players:

```{r}

batters |> 
  arrange(desc(performance))

```

You can find a good explanation of this problem and how to overcome it at http://varianceexplained.org/r/empirical_bayes_baseball/ and https://www.evanmiller.org/how-not-to-sort-by-average-rating.html
