---
title: "5 Data tidying"
format: pdf
editor: visual
---

### **5.1 Introduction**

“Happy families are all alike; every unhappy family is unhappy in its own way.” *— Leo Tolstoy*

“Tidy datasets are all alike, but every messy dataset is messy in its own way.” *— Hadley Wickham*


```{r}
#| message: false
#| warning: false

library(tidyverse)

```

### **5.2 Tidy data**

There are three interrelated rules that make a dataset tidy:

1. Each variable is a column; each column is a variable.
2. Each observation is a row; each row is an observation.
3. Each value is a cell; each cell is a single value.

This is Codd’s 3rd normal form, but with the constraints framed in statistical language, and the focus put on a single dataset rather than the many connected datasets common in relational databases. Messy data is any other arrangement of the data.

https://tidyr.tidyverse.org/

For a given dataset, it’s usually easy to figure out what are observations and what are variables, but it is surprisingly difficult to precisely define variables and observations in general. For example, if the columns in the classroom data were height and weight we would have been happy to call them variables. If the columns were height and width, it would be less clear cut, as we might think of height and width as values of a dimension variable. If the columns were home phone and work phone, we could treat these as two variables, but in a fraud detection environment we might want variables phone number and number type because the use of one phone number for multiple people might suggest fraud. A general rule of thumb is that it is easier to describe functional relationships between variables (e.g., z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of observations (e.g., average of group a vs. average of group b) than between groups of columns.

Packages in the tidyverse are designed to work with tidy data:

```{r}

# Compute rate per 10,000
table1 |>
  mutate(rate = cases / population * 10000)

# Compute total cases per year
table1 |> 
  group_by(year) |> 
  summarize(total_cases = sum(cases))

# Visualize changes over time
ggplot(
  table1, 
  aes(x = year, y = cases)
) +
  geom_line(
    aes(group = country), 
    color = "grey50"
  ) +
  geom_point(
    aes(color = country, shape = country)
  ) +
  # x-axis breaks at 1999 and 2000
  scale_x_continuous(breaks = c(1999, 2000)) 

```
### **Exercises**

**1. For each of the sample tables, describe what each observation and each column represents.**

In table1, each row is one country–year observation. It tells, for a given country and year, how many cases of a disease were recorded and what the total population was.

Columns:

  - country: Country name (e.g. Afghanistan, Brazil).
  
  - year: Calendar year of data collection.

  - cases: Number of people with the disease in that country and year.

  - population: Total number of people living in that country in that year.

Conceptually, table1 is already tidy: one row per country–year, one column per variable.

```{r}

table1

```

In table2, each row is still one country–year–type observation, but the disease and population counts are stacked into a longer format.

Columns:

  - country: Country name, same meaning as in table1.
  
  - year: Calendar year, same meaning as in table1.
  
  - type: Indicates which kind of count the row refers to, either "cases" or "population".
  
  - count: The numeric value corresponding to type (number of disease cases if type == "cases", or total population if type == "population").

Here, the underlying variables are “cases” and “population”, but they are represented by the pair (type, count) rather than separate columns.

```{r}

table2

```
In table3, each row is again one country–year observation, but the disease information is expressed as a rate instead of separate counts.

Columns:

  - country: Country name, same as above.
  
  - year: Calendar year, same as above.
  
  - rate: Disease rate, defined conceptually as cases divided by population for that country and year (often written in a “cases / population” style in the example data).

```{r}

table3

```

So, across table1, table2, and table3, country and year always define the observation; what changes is whether the absolute counts are stored in separate columns, as long format type–value pairs, or as a single rate variable.


**2. Sketch out the process you’d use to calculate the rate for table2 and table3. You will need to perform four operations:**

  - Extract the number of TB cases per country per year.
  - Extract the matching population per country per year.
  - Divide cases by population, and multiply by 10000.
  - Store back in the appropriate place.

**table2**

table2 has columns country, year, type (either "cases" or "population"), and count. Conceptually you want to “widen” it so you get separate cases and population columns, then compute the rate.

1. Extract TB cases per country–year

    - Look at all rows where type is "cases".
    
    - For each country and year combination, take the count value from these rows as the number of TB cases.
    
    - Mentally, you can imagine building a temporary list or table: (country, year, cases).

2. Extract matching population per country–year

    - Now look at all rows where type is "population".
    
    - For each country and year, take the count value as the total population.
    
    - Conceptually you have a second list: (country, year, population).
    
    - Match these two lists together by country and year so that, for every country–year, you have both cases and population side by side.

3. Compute rate = cases / population * 10,000

    - For each matched country–year record, divide the number of cases by the population to get a proportion.
    
    - Multiply that proportion by 10,000 to convert it to “cases per 10,000 people.”
    
    - Store that numeric result as a new value called rate for that country–year.
  
4. Store back in the appropriate place

    - Decide on the structure you want going forward. Typically, you’d create a table with columns country, year, cases, population, and rate.
    
    - Conceptually, this is “table2 reshaped”: one row per country–year, the new rate column added there.
    
    - If you need to keep the original long format for some reason, you’d treat this as a derived table, but the core idea is that rate lives alongside the corresponding cases and population for each country–year.

**table3**

table3 has columns country, year, and rate, but here rate is stored as a textual composite like “cases/population” rather than a numeric rate per 10,000.

1. Extract TB cases per country–year

    - Start from the composite rate value in each row (e.g. “745/19987071”).
    
    - Conceptually split that value into two pieces separated by “/”: the first part is the number of cases, the second part is the population.
    
    - Treat the first piece as the TB cases count for that row’s country–year, forming (country, year, cases).
  
2. Extract matching population per country–year

    - For the same row, take the second piece of the split string as the population.
    
    - Convert it to a numeric population value and pair it with the cases for that country–year, i.e. (country, year, cases, population).

3. Compute rate = cases / population * 10,000

    - For each row, divide the numeric cases by the numeric population to get a proportion.
    
    - Multiply that proportion by 10,000 to express it as “cases per 10,000 people.”
    
    - This gives you a proper numeric rate variable derived from the decomposed cases and population.

4. Store back in the appropriate place

    - Replace or augment the original rate column with your newly computed numeric rate per 10,000.
    
    - Keep cases and population as their own columns so the table now looks like: country, year, cases, population, rate.
    
    - At this point, table3 is structurally similar to the reshaped version of table2, and you can use it in the same way for analysis or visualization.

### **5.3 Lengthening data**

#### Data in column names

The billboard dataset records the billboard rank of songs in the year 2000:

```{r}

billboard

```
In this dataset, each observation is a song. The first three columns (artist, track and date.entered) are variables that describe the song. Then we have 76 columns (wk1-wk76) that describe the rank of the song in each week1. Here, the column names are one variable (the week) and the cell values are another (the rank).

To tidy this data, we’ll use pivot_longer():

```{r}

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )

```
After the data, there are three key arguments:

- **cols** specifies which columns need to be pivoted, i.e. which columns aren’t variables. This argument uses the same syntax as select() so here we could use !c(artist, track, date.entered) or starts_with("wk").

- **names_to names** the variable stored in the column names, we named that variable week.

- **values_to names** the variable stored in the cell values, we named that variable rank.

Note that in the code "week" and "rank" are quoted because those are new variables we’re creating, they don’t yet exist in the data when we run the pivot_longer() call.

Now let’s turn our attention to the resulting, longer data frame. What happens if a song is in the top 100 for less than 76 weeks? Take 2 Pac’s “Baby Don’t Cry”, for example. The above output suggests that it was only in the top 100 for 7 weeks, and all the remaining weeks are filled in with missing values. These NAs don’t really represent unknown observations; they were forced to exist by the structure of the dataset2, so we can ask pivot_longer() to get rid of them by setting values_drop_na = TRUE:

```{r}

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )

```

This data is now tidy, but we could make future computation a bit easier by converting values of week from character strings to numbers using mutate() and readr::parse_number(). parse_number() is a handy function that will extract the first number from a string, ignoring all other text.

```{r}

billboard_longer <- 
  billboard |> 
    pivot_longer(
      cols = starts_with("wk"), 
      names_to = "week", 
      values_to = "rank",
      values_drop_na = TRUE
    ) |> 
    mutate(
      week = parse_number(week)
    )

billboard_longer

```

How song ranks vary over time

```{r}
billboard_longer |> 
  ggplot(
    aes(x = week, y = rank, group = track)
  ) + 
  geom_line(alpha = 0.25) + 
  scale_y_reverse()

```

### **5.3.2 How does pivoting work?**

Suppose we have three patients with ids A, B, and C, and we take two blood pressure measurements on each patient. 

```{r}

df <- 
  tribble(
    ~id,  ~bp1, ~bp2,
     "A",  100,  120,
     "B",  140,  115,
     "C",  120,  125
  )

```

We want our new dataset to have three variables: id (already exists), measurement (the column names), and value (the cell values). To achieve this, we need to pivot df longer:

```{r}

df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )

```

### **5.3.3 Many variables in column names**

A more challenging situation occurs when you have multiple pieces of information crammed into the column names, and you would like to store these in separate new variables. For example, take the who2 dataset, the source of table1 and friends that you saw above. 

This dataset, collected by the World Health Organisation, records information about tuberculosis diagnoses. There are two columns that are already variables and are easy to interpret: country and year. They are followed by 56 columns like sp_m_014, ep_m_4554, and rel_m_3544. If you stare at these columns for long enough, you’ll notice there’s a pattern. Each column name is made up of three pieces separated by _. The first piece, sp/rel/ep, describes the method used for the diagnosis, the second piece, m/f is the gender (coded as a binary variable in this dataset), and the third piece, 014/1524/2534/3544/4554/5564/65 is the age range (014 represents 0-14, for example).

So in this case we have six pieces of information recorded in who2: the country and the year (already columns); the method of diagnosis, the gender category, and the age range category (contained in the other column names); and the count of patients in that category (cell values):

```{r}

who2

```
This is only a minor variation on the simpler case above. Instead of the column names pivoting into a single column, they pivot into multiple columns.

```{r}

who2 |> 
  pivot_longer(
    cols = !(country:year),
    names_to = c("diagnosis", "gender", "age"), 
    names_sep = "_",
    values_to = "count"
  )

```

### **5.3.4 Data and variable names in the column headers**

Column names include a mix of variable values and variable names. 

This dataset contains data about five families, with the names and dates of birth of up to two children. The new challenge in this dataset is that the column names contain the names of two variables (dob, name) and the values of another (child, with values 1 or 2). 


```{r}

household

```

To solve this problem we again need to supply a vector to names_to but this time we use the special ".value" sentinel; this isn’t the name of a variable but a unique value that tells pivot_longer() to do something different. This overrides the usual values_to argument to use the first component of the pivoted column name as a variable name in the output. 

When you use ".value" in names_to, the column names in the input contribute to both values and variable names in the output:

```{r}

household |> 
  pivot_longer(
    cols = !family, 
    names_to = c(".value", "child"), 
    names_sep = "_", 
    values_drop_na = TRUE
  )

```

### **5.4 Widening data**

pivot_wider() makes datasets wider by increasing columns and reducing rows and helps when one observation is spread across multiple rows. This seems to arise less commonly in the wild, but it does seem to crop up a lot when dealing with governmental data.

cms_patient_experience, is a dataset from the Centers of Medicare and Medicaid services that collects data about patient experiences:

```{r}

cms_patient_experience

```

The core unit being studied is an organization, but each organization is spread across six rows, with one row for each measurement taken in the survey organization. We can see the complete set of values for measure_cd and measure_title by using distinct():


```{r}

cms_patient_experience |> 
  distinct(measure_cd, measure_title)

```

Neither of these columns will make particularly great variable names: measure_cd doesn’t hint at the meaning of the variable and measure_title is a long sentence containing spaces. We’ll use measure_cd as the source for our new column names for now, but in a real analysis you might want to create your own variable names that are both short and meaningful.

pivot_wider() has the opposite interface to pivot_longer(): instead of choosing new column names, we need to provide the existing columns that define the values (values_from) and the column name (names_from):


```{r}

cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )


```

We still seem to have multiple rows for each organization. That’s because, we also need to tell pivot_wider() which column or columns have values that uniquely identify each row; in this case those are the variables starting with "org":


```{r}

cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )

```

### **5.4.1 How does pivot_wider() work?**

We have a very simple dataset with two patients with ids A and B. We have three blood pressure measurements on patient A and two on patient B:

```{r}

df <- 
  tribble(
    ~id, ~measurement, ~value,
    "A",        "bp1",    100,
    "B",        "bp1",    140,
    "B",        "bp2",    115, 
    "A",        "bp2",    120,
    "A",        "bp3",    105
  )

```

We’ll take the values from the value column and the names from the measurement column:


```{r}

df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )

```

To begin the process pivot_wider() needs to first figure out what will go in the rows and columns. The new column names will be the unique values of measurement.


```{r}

df |> 
  distinct(measurement) |> 
  pull()

```

By default, the rows in the output are determined by all the variables that aren’t going into the new names or values. These are called the id_cols. Here there is only one column, but in general there can be any number.


```{r}

df |> 
  select(!measurement & !value) |> 
  distinct()

```

pivot_wider() then combines these results to generate an empty data frame:

```{r}

df |> 
  select(!measurement & !value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)

```

It then fills in all the missing values using the data in the input.

What happens if there are multiple rows in the input that correspond to one cell in the output. The example below has two rows that correspond to id “A” and measurement “bp1”:

```{r}

df <- 
  tribble(
    ~id, ~measurement, ~value,
    "A",        "bp1",    100,
    "A",        "bp1",    102,
    "A",        "bp2",    120,
    "B",        "bp1",    140, 
    "B",        "bp2",    115
  )

```

If we attempt to pivot this we get an output that contains list-columns:

```{r}

df |>
  pivot_wider(
    names_from = measurement,
    values_from = value
  )

```
Since you don’t know how to work with this sort of data yet, you’ll want to follow the hint in the warning to figure out where the problem is:

```{r}

df |> 
  group_by(id, measurement) |> 
  summarize(n = n(), .groups = "drop") |> 
  filter(n > 1)

```

It’s then up to you to figure out what’s gone wrong with your data and either repair the underlying damage or use your grouping and summarizing skills to ensure that each combination of row and column values only has a single row.







